The Form of Uncertainty Affects Selection for Social Learning

### 2023-02-25 ms.tex
updated supplement figures labels to match new main text figure labels

\author[1,2,*]{Matthew A. Turner}
\author[3,4]{Cristina Moya}
\author[5,6,7]{Paul E. Smaldino}
\author[1,2,7]{James~Holland~Jones}

\affil[1]{Department of Earth System Science, Stanford University, Stanford, CA 94305 USA}
\affil[2]{Division of Social Sciences, Stanford Doerr School of Sustainability, Stanford University, Stanford, CA 94305 USA}
\affil[3]{Department of Anthropology, University of California at Davis, Davis, CA 95616 USA}
\affil[4]{Centre for Culture and Evolution, Brunel University London, Uxbridge, UB8 3PH UK}
\affil[5]{Cognitive and Information Sciences, University of California at Merced, Merced, CA 95340 USA}
\affil[6]{Santa Fe Institute, Santa Fe, NM 87501 USA}
\affil[7]{Center for Advanced Study in the Behavioral Sciences, Stanford University, Stanford, CA 94305 USA}
\affil[*]{Corresponding author; email: \href{mailto:maturner@stanford.edu}{maturner@stanford.edu}}

@@

\begin{figure}
  \caption{As the selection-set size, $B$, and effective lifespan, $L$, increase (set equal along x-axis), 
  social learning frequently evolves across an increasingly large range of
  environmental variability values, i.e., up to
  increasingly large values of the social learning ceiling
(y-axis; defined in Equation~\ref{eq:slCeiling}).}
  \label{fig:slCeiling}
  \centering
    \includegraphics[width=0.8\textwidth]{Figures/SL_Extinction_Variability.pdf}
\end{figure}

@@

When payoff ambiguity and selection-set size was small ($\pilow=0.1$ and $B=2$), and lifespan was long ($L \geq 4$), asocial learning fixated as a hedge against periods of protracted environmental instability, even though expected
homogeneous social payoffs were greater than homogeneous asocial payoffs (Figure~\ref{fig:payoffTimeseries}A). Runs of environmental change strongly affected the geometric mean fitness of social learners who were weighing
outdated information too heavily. \edit{In other words, environmental change lowers social learners' mean payoffs, but social learner payoffs build back up when (blue dots, Figure~\ref{fig:payoffTimeseries}A).  Even though an homogenous population of social learners would outperform individual learners in this case, unpredictable environmental stochasticity undermines the reliability of social learning.  When lifespan was minimal ($L=1$) and environmental variability was moderate ($u\approx0.5$), the opposite, unexpected outcome could occur where social learning fixates despite the fact that an homogenous population of asocial learners would outperform an homogenous population of social learners (Figure~\ref{fig:payoffTimeseries}B). In this case, it takes relatively few consecutive generations with environmental stability to attract the population to all become social learners.}

@@

Our model necessarily made simplifying assumptions. These were chosen judiciously to
address the main question of how uncertainty affects the evolution of social
learning. Two critical assumptions we made have important consequences for comparing
our results with previous findings in the literature. First, we assumed all agents
engage in individual learning. This means that individual learning is effectively
free, though potentially inaccurate. Although producing new knowledge may be costly,
many other adaptive problems take the form of essentially cost-free individual
learning. For example, it is hard not to encode association between clouds and rain,
between thirst and drinking water, or between friendship and warm laughter.
\edit{If there were an exogenous cost of individual learning, this would likely
  increase the social learning ceiling across uncertainty parameter settings.
Our model operationalizes environmental variability as fully stochastic,
which forces evolution towards either a population of social or asocial learners
since environmental change is unpredictable, unlike precursor
models that allowed for persistent frequency
dependence~\citep{BoydRicherson1985,Rogers1988,Feldman1996}.} Second, we only allow for
intergenerational social learning and environmental change. Within generations,
agents can only learn individually. This assumption allows us to replicate the
classic finding that if the environment changes too quickly, socially-learned
information may become \edit{outdated. Adding horizontal social learning between
vertical/oblique learning in this model would likely expand the range of
environmental variability over which social learning would
evolve~\citep{Turner2022}, i.e., horizontal learning would increase what we have
called the social learning ceiling. However, depending on the
setting, there may be little difference between horizontal and oblique learning, for
example if the generations are purely cultural ‘generations’ where parents, and
others of the parent's generation, may also be peers.}

We could have made alternative choices for a range of other structural features, the future exploration of which
will be important to the development of more parsimonious theories of social learning.  
For example, we only considered success-biased learning, although people may use conformity or other context biases to socially learn~\citep{BoydRicherson1985,Muthukrishna2016a,Smaldino2018b}. This modeling
choice maximizes the potential benefit to social learners, which suggests that
other mechanisms, like conformity, may lead to \edit{weaker selection} overall. 
\edit{Furthermore, our learning procedure could be interpreted as direct copying of
latent psychological variables (expected payoffs for each behavior); if children
learned instead from observations of the parent's behaviors without knowing rewards,
which in some cases would be a more realistic assumption~\citep{Wu2022b}, this may weaken the strength of selection for social learning.} We
also assumed that the number of behaviors and their payoffs were constant within a
given simulation, but this fails to account for evolutionary feedback which creates
new behavioral opportunities as time progresses\edit{~\citep{Chimento2022}}, e.g., via niche
construction~\citep{Smaldino2012a,Heras-Escribano2020} or cumulative cultural
evolution~\citep{Smolla2019,Derex2020}.  Group structure and processes such as
homophily and other group-level biases could inhibit the evolution of social
learning since successful out-group teachers could be cut off from in-group
learners~\citep{Golub2012}. We assumed agents choose behaviors via a softmax
\edit{decision rule, which is more cognitively sophisticated than most
social-learning models and captures the adaptiveness of real human learning
mechanisms. This modeling decision was essential for developing our understanding of
how social and individual learning can combine to multiply problem solving capacity
of our agents, though it is still a pale shadow of the true sophistication of human
learning capabilities~\citep{Schulz2020a,Wu2022}}. More realistic and powerful individual learning could further support the evolution of
social learning. These considerations provide novel opportunities to expand the evolutionary theory of social learning by adding
such learning mechanisms to our model.

@@

%We found that the 
The evolution of social learning is a complex phenomenon, dependent on many interrelated factors, yet it is widespread across a
range of taxa. Although social learning is broadly studied, few studies have taken a systematic approach to understanding the role
of uncertainty in the evolution \edit{of} social learning. Understanding the role of uncertainty in social evolution is particularly important
for understanding human evolution as evidence increasingly suggests that the genus \emph{Homo} emerged in the context of substantial
(nonstationary) environmental variability \citep{anton_etal2014, levin2015}, and our global dispersal
suggests that human groups regularly face novel environments. \edit{In our model, more numerous behavioral options meant agents more flexibly adjusted their behavior if social learning provided outdated information since in the softmax decision rule, more options means relatively less weight is assigned to the most profitable behavior known to agents, and greater weight is put on unexplored behaviors or those behaviors perceived to be less optimal. While Rogers'-style models of social learning evolution may also predict greater social learning reliance in more challenging contexts (larger selection-set sizes), they cannot predict that social learning should be favored for such high values of
environmental variability as we observed here.} 
By carefully identifying and operationalizing different forms of uncertainty common in social-learning models, we developed a more thorough understanding of how uncertainty can favor social learning and how different forms of uncertainty interact to favor its evolution. By explicitly modeling a key individual-level learning mechanism shared across taxa, namely the ability to adjust behavior to uncertainty and new information, we found that social learning could evolve even if it provides outdated information. %Social learning is a critical component of problem solving among social creatures. Our work identified and systematized some important forms of uncertainty and catalogued their effects on the evolution of social learning.
Our work, then, both advances the theory of social learning and could eventually be of practical use in identifying specific mechanisms underlying adaptations to uncertain environments.

@@

\begin{figure}
  \centering
  % \addtocounter{figure}{-1}
  \caption{
	Sensitivity analysis of the main results for two population
	sizes, $N=50,200$. Recall $N=1000$ was used to generate main 
	text results.
  }
  \label{fig:populationSensitivity}
  \begin{subfigure}{\textwidth}
	\caption{$N=50$}
	\includegraphics[width=\textwidth]{Figures/supplement/numagents=50/mainResultsPlots.pdf}
  \end{subfigure}
\end{figure}

\begin{figure}
  \ContinuedFloat
	\begin{subfigure}{\textwidth}
	  \caption{$N=200$}
	  \includegraphics[width=\textwidth]{Figures/supplement/numagents=200/mainResultsPlots.pdf}
	\end{subfigure}
\end{figure}

@@

\vspace{-3em}
\begin{figure}
  \centering
   % \addtocounter{figure}{-1}
  \caption{Number of prospective teachers sensitivity analysis for $N_T=2,20$. Recall
  $N_T=5$ was used to generate main text results.}
  \label{fig:nteachersSensitivity}
  \vspace{2em}
  \begin{subfigure}{\textwidth}
	\caption{$N_T = 2$}
	\includegraphics[width=\textwidth]{Figures/supplement/nteachers=2/mainResultsPlots.pdf}
  \end{subfigure}
\end{figure}
% \newpage
% \begin{figure}
%   \ContinuedFloat
%   \begin{subfigure}{\textwidth}
% 	\caption{$N_T = 10$}
% 	\includegraphics[width=\textwidth]{Figures/supplement/nteachers=10/mainResultsPlots.pdf}
%   \end{subfigure}
% \end{figure}
\newpage
\begin{figure}
  \ContinuedFloat
  \begin{subfigure}{\textwidth}
	\caption{$N_T = 20$}
	\includegraphics[width=\textwidth]{Figures/supplement/nteachers=20/mainResultsPlots.pdf}
  \end{subfigure}
\end{figure}

@@

\vspace{-3em} \begin{figure} %\addtocounter{figure}{-1} 
  \centering
  % \addtocounter{figure}{-1}
  \caption{Sensitivity analysis of the main results for the softmax parameter $\beta = 100$ and
  $\beta=1$. Recall the main results were obtained with $\beta = 10$.}
  \label{fig:softmaxSensitivity} \vspace{2em}
  \begin{subfigure}{\textwidth}
	\caption{$\beta = 1$}
	\includegraphics[width=\textwidth]{Figures/supplement/sensitivity_tau=1.0/mainResultsPlots.pdf}
  \end{subfigure}
\end{figure}

@@

\begin{figure}
  \ContinuedFloat
  \begin{subfigure}{\textwidth}
	\caption{$\beta = 100$}
	\includegraphics[width=\textwidth]{Figures/supplement/sensitivity_tau=0.01/mainResultsPlots.pdf}
  \end{subfigure}
\end{figure}

>>>>

\author[1,2,*]{Matthew A. Turner}
\author[3]{Cristina Moya}
\author[4,5,6]{Paul E. Smaldino}
\author[1,2,6]{James~Holland~Jones}

\affil[1]{Department of Earth System Science, Stanford University, Stanford, CA 94305 USA}
\affil[2]{Division of Social Sciences, Stanford Doerr School of Sustainability, Stanford University, Stanford, CA 94305 USA}
\affil[3]{Department of Anthropology, University of California at Davis, Davis, CA 95616 USA}
\affil[4]{Cognitive and Information Sciences, University of California at Merced, Merced, CA 95340 USA}
\affil[5]{Santa Fe Institute, Santa Fe, NM 87501 USA}
\affil[6]{Center for Advanced Study in the Behavioral Sciences, Stanford University, Stanford, CA 94305 USA}
\affil[*]{Corresponding author; email: \href{mailto:maturner@stanford.edu}{maturner@stanford.edu}}

@@

\begin{figure}
  \caption{As the selection-set size, $B$, and effective lifespan, $L$, increase (set equal along x-axis), 
  social learning frequently evolves across an increasingly large range of
  environmental variability values, i.e., up to
  increasingly large values of the social learning ceiling
(y-axis; defined in Equation~\ref{eq:slCeiling}).}
  \label{fig:slCeiling}
  \centering
    \includegraphics[width=0.8\textwidth]{Figures/SL_ceiling.pdf}
\end{figure}

@@

When payoff ambiguity and selection-set size was small ($\pilow=0.1$ and $B=2$), and lifespan was long ($L \geq 4$), asocial learning fixated as a hedge against periods of protracted environmental instability, even though expected
homogeneous social payoffs were greater than homogeneous asocial payoffs (Figure~\ref{fig:payoffTimeseries}A). Runs of environmental change strongly affected the geometric mean fitness of social learners who were weighing
outdated information too heavily. \edit{In other words, environmental change lowers social learners' mean payoffs, but social learner payoffs build back up as the information quality in the population improves (blue dots, Figure~\ref{fig:payoffTimeseries}A).  Even though an homogenous population of social learners would outperform individual learners in this case, unpredictable environmental stochasticity undermines the reliability of social learning.  When lifespan was minimal ($L=1$) and environmental variability was moderate ($u\approx0.5$), the opposite, unexpected outcome could occur where social learning fixates despite the fact that an homogenous population of asocial learners would outperform an homogenous population of social learners (Figure~\ref{fig:payoffTimeseries}B). In this case, it takes relatively few consecutive generations with environmental stability to attract the population to all become social learners.}

@@

Our model necessarily made simplifying assumptions. These were chosen judiciously to
address the main question of how uncertainty affects the evolution of social
learning. Two critical assumptions we made have important consequences for comparing
our results with previous findings in the literature. First, we assumed all agents
engage in individual learning. This means that individual learning is effectively
free, though potentially inaccurate. Although producing new knowledge may be costly,
many other adaptive problems take the form of essentially cost-free individual
learning. For example, it is hard not to encode association between clouds and rain,
between thirst and drinking water, or between friendship and warm laughter.
\edit{If there were an exogenous cost of individual learning, this would likely
  increase the social learning ceiling across uncertainty parameter settings.
Our model operationalizes environmental variability as fully stochastic,
which forces evolution towards either a population of social or asocial learners
since environmental change is unpredictable, unlike precursor
models that allowed for persistent frequency
dependence~\citep{BoydRicherson1985,Rogers1988,Feldman1996}.} Second, we only allow for
intergenerational social learning and environmental change. Within generations,
agents can only learn individually. This assumption \edit{is consistent with previous modeling work starting at least from \cite{Rogers1988}, and therefore} allows us to replicate the
classic finding that if the environment changes too quickly, socially-learned
information may become \edit{outdated. Adding horizontal social learning between
vertical/oblique learning in this model would likely expand the range of
environmental variability over which social learning would
evolve~\citep{Turner2022}, i.e., horizontal learning would increase what we have
called the social learning ceiling. However, depending on the
setting, there may be little difference between horizontal and oblique learning, for
example if the generations are purely cultural ‘generations’ where parents, and
others of the parent's generation, may also be peers.}

We could have made alternative choices for a range of other structural features, the future exploration of which
will be important to the development of more parsimonious theories of social learning.  
For example, we only considered success-biased learning, although people may use conformity or other context biases to socially learn~\citep{BoydRicherson1985,Muthukrishna2016a,Smaldino2018b}. This modeling
choice maximizes the potential benefit to social learners, which suggests that
other mechanisms, like conformity, may lead to \edit{weaker selection} overall. 
\edit{Furthermore, our learning procedure could be interpreted as direct copying of latent psychological variables (expected payoffs for each behavior). Such detailed information is most likely transmitted through teaching wherein the model explicitly informs the learner of their experiences. If children learned instead from observations of the parent's behaviors without knowing rewards, which in some cases would be a more realistic assumption~\citep{Wu2022b}, this may weaken the strength of selection for social learning.} We also assumed that the number of behaviors and their payoffs were constant within a
given simulation, but this fails to account for evolutionary feedback which creates
new behavioral opportunities as time progresses\edit{~\citep{Chimento2022}}, e.g., via niche
construction~\citep{Smaldino2012a,Heras-Escribano2020} or cumulative cultural
evolution~\citep{Smolla2019,Derex2020}.  Group structure and processes such as
homophily and other group-level biases could inhibit the evolution of social
learning since successful out-group teachers could be cut off from in-group
learners~\citep{Golub2012}. We assumed agents choose behaviors via a softmax
\edit{decision rule, which is more cognitively sophisticated than most
social-learning models and captures the adaptiveness of real human learning
mechanisms. This modeling decision was essential for developing our understanding of
how social and individual learning can combine to multiply problem solving capacity
of our agents, though it is still a pale shadow of the true sophistication of human
learning capabilities~\citep{Schulz2020a,Wu2022}}. More realistic and powerful individual learning could further support the evolution of
social learning. These considerations provide novel opportunities to expand the evolutionary theory of social learning by adding
such learning mechanisms to our model.

@@

%We found that the 
The evolution of social learning is a complex phenomenon, dependent on many interrelated factors, yet it is widespread across a
range of taxa. Although social learning is broadly studied, few studies have taken a systematic approach to understanding the role
of uncertainty in the evolution \edit{of} social learning. Understanding the role of uncertainty in social evolution is particularly important
for understanding human evolution as evidence increasingly suggests that the genus \emph{Homo} emerged in the context of substantial
(nonstationary) environmental variability \citep{anton_etal2014, levin2015}, and our global dispersal
suggests that human groups regularly face novel environments. \edit{In our model, more behavioral options meant agents more flexibly adjusted their behavior if social learning provided outdated information since in the softmax decision rule, more options means relatively less weight is assigned to the most profitable behavior known to agents, and greater weight is put on unexplored behaviors or those behaviors perceived to be less optimal. While Rogers'-style models of social learning evolution may also predict greater social learning reliance in more challenging contexts (larger selection-set sizes), they would not predict that social learning should be favored for such high values of
environmental variability as we observed here.} 
By carefully identifying and operationalizing different forms of uncertainty common in social-learning models, we developed a more thorough understanding of how uncertainty can favor social learning and how different forms of uncertainty interact to favor its evolution. By explicitly modeling a key individual-level learning mechanism shared across taxa, namely the ability to adjust behavior to uncertainty and new information, we found that social learning could evolve even if it provides outdated information. %Social learning is a critical component of problem solving among social creatures. Our work identified and systematized some important forms of uncertainty and catalogued their effects on the evolution of social learning.
Our work, then, both advances the theory of social learning and could eventually be of practical use in identifying specific mechanisms underlying adaptations to uncertain environments.

@@

\begin{figure}
  \centering
  % \addtocounter{figure}{-1}
  \caption{
	Sensitivity analysis of the main results for two population
	sizes, $N=50,200$. Recall $N=1000$ was used to generate main 
	text results.
  }
  \label{fig:populationSensitivity}
  \begin{subfigure}{\textwidth}
	\caption{Population size, $N=50$}
	\includegraphics[width=\textwidth]{Figures/supplement/numagents=50/mainResultsPlots.pdf}
  \end{subfigure}
\end{figure}

\begin{figure}
  \ContinuedFloat
	\begin{subfigure}{\textwidth}
	  \caption{Population size, $N=200$}
	  \includegraphics[width=\textwidth]{Figures/supplement/numagents=200/mainResultsPlots.pdf}
	\end{subfigure}
\end{figure}

@@

\vspace{-3em}
\begin{figure}
  \centering
   % \addtocounter{figure}{-1}
  \caption{Number of prospective teachers sensitivity analysis for $N_T=2,20$. Recall
  $N_T=5$ was used to generate main text results.}
  \label{fig:nteachersSensitivity}
  \vspace{2em}
  \begin{subfigure}{\textwidth}
	\caption{Number of prospective teachers, $N_T = 2$}
	\includegraphics[width=\textwidth]{Figures/supplement/nteachers=2/mainResultsPlots.pdf}
  \end{subfigure}
\end{figure}
% \newpage
% \begin{figure}
%   \ContinuedFloat
%   \begin{subfigure}{\textwidth}
% 	\caption{$N_T = 10$}
% 	\includegraphics[width=\textwidth]{Figures/supplement/nteachers=10/mainResultsPlots.pdf}
%   \end{subfigure}
% \end{figure}
\newpage
\begin{figure}
  \ContinuedFloat
  \begin{subfigure}{\textwidth}
	\caption{Number of prospective teachers, $N_T = 20$}
	\includegraphics[width=\textwidth]{Figures/supplement/nteachers=20/mainResultsPlots.pdf}
  \end{subfigure}
\end{figure}

@@

\vspace{-3em} \begin{figure} %\addtocounter{figure}{-1} 
  \centering
  % \addtocounter{figure}{-1}
  \caption{Sensitivity analysis of the main results for the softmax parameter $\beta = 100$ and
  $\beta=1$. Recall the main results were obtained with $\beta = 10$.}
  \label{fig:softmaxSensitivity} \vspace{2em}
  \begin{subfigure}{\textwidth}
	\caption{Softmax greediness, $\beta = 1$}
	\includegraphics[width=\textwidth]{Figures/supplement/sensitivity_tau=1.0/mainResultsPlots.pdf}
  \end{subfigure}
\end{figure}

@@

\begin{figure}
  \ContinuedFloat
  \begin{subfigure}{\textwidth}
	\caption{Softmax greediness, $\beta = 100$}
	\includegraphics[width=\textwidth]{Figures/supplement/sensitivity_tau=0.01/mainResultsPlots.pdf}
  \end{subfigure}
\end{figure}
